\documentclass[11pt]{article}

\usepackage{amsmath,amsthm}
\usepackage[headings]{fullpage}
\usepackage[utopia]{mathdesign}
\usepackage{color}

\pagestyle{myheadings}
\markboth{Matrix conditioning}{}

\input{../../fncextra}

\begin{document}

\begin{center}
  \bf Terms and conditions
\end{center}

Previously you have seen how to blur an image, represented as an
$m\times n$ pixel intensity matrix $\mX$, through multiplying by a matrix on each side:
\begin{equation}
  \label{eq:2}
   \mZ=\mV\, \mX \, \mH,
\end{equation}
where $\mV = (\mB_m)^k$, $\mH = (\mB_n)^k$, 
$B_i$ is the $i\times i$ matrix made by your function
\texttt{blurmatrix}, and $k$ is a positive integer. You have also seen that deblurring can be accomplished by multiplying by matrix inverses on each side of $\mZ$ (as computed equivalently by solving linear systems). However, while the restoration seems perfect for $k=1$, it fails completely for larger $k$.

Matrix condition numbers explain these observations. The blurred matrix $\mZ$ is perturbed by an amount comparable to machine precision. For $k=1$, the condition numbers of $\mV$ and $\mH$ are not large enough to amplify this error up to the same order of magnitude as $\mX$ itself, so the noise is not perceived. But at some $k>1$, the condition numbers are so large that the noise is amplified enough to overwhelm the expected result.

Let $\bfx$ be a single column of $\mX$. For the case of vertical blurring only, we have $\bfz=\mV\bfx$ as a column of $\mZ$. We solve $\mV \bfy = \bfz$ for $\bfy$, which is mathematically the same as $\bfx$. Due to machine precision, though, the perturbation to $\bfz$ causes an error satisfying
\begin{equation}
	\label{bound}
	\frac{\|\bfy-\bfx\|}{\|\bfx\|} \le \kappa(\mV)\macheps,
\end{equation}
with $\kappa$ being the matrix condition number. 

\subsection*{Preparation}

Read Section~2.8. Make sure your working \texttt{blurmatrix.m} is available. 

\subsection*{Goals}

You will compute condition numbers of blur matrix powers and compare them to the errors of repeated blur/deblur operations. 

\subsection*{Procedure}

\begin{enumerate}
\item The condition number of a matrix has two factors, $\|\mA\|$ and $\|\mA^{-1}\|$. First you will show that the $\|\mA\|$ term makes no trouble. For $n=50,100,150,\ldots,800$, plot $\|\mB_n\|$ versus $n$. 

\item For the same $n$ as in step 1, plot $\kappa(\mB_n)$ versus $n$. You should use a log-log scale for this graph and get essentially a straight line. This implies that $\log \kappa \approx a \log n + b$, or $\kappa \approx C n^p$. 

\item Let $\mV=\mB_{100}$. For $k=1,2,\ldots,8$, plot $\kappa(\mV^k)$ as a function of $k$. This time the graph is straight on a semi-log scale, which implies $\kappa \approx C q^k$. 

\item Let $\bfx$ be a random vector of length 100. For $k=1,2,\ldots,8$, let $\bfz=\mV\bfx$ and then solve $\mV\bfy=\bfz$ for $\bfy$. Record the relative error in the result. Then make a table showing both sides of the inequality~\eqref{bound}.

\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
