\documentclass[11pt]{article}

\usepackage[headings]{fullpage}
\usepackage[utopia]{mathdesign}

\pagestyle{myheadings}
\markboth{MATH426/CISC410}{MATH426/CISC410}

\input{../../fncextra}

\begin{document}

\begin{center}
  \bf Not entirely stable
\end{center}

Consider the function
\[
f(x) = \frac{e^x-1}{x}.
\]
A computation of its relative condition number $\kappa$ shows that $|\kappa(x)|<1$ for all $|x|<1$. In that sense, $f$ is ``easy'' to compute accurately. In practice, though, it's not so simple.

An obvious sequence of steps to compute $f$ is as follows:
\begin{equation}
  \label{direct}
  y_1 = e^x, \qquad y_2 = y_1 - 1, \qquad y_3 = y_2/x.
\end{equation}
The operations for $y_1$ and $y_3$ are well conditioned for $|x|<1$, but the subtraction to get $y_2$ will suffer from cancellation error if $y_1\approx 1$, or $x\approx 0$. That error makes this sequence of operations unstable for $x\approx 0$.

Now consider a power series expansion of $f$,
\begin{equation}
  \label{series}
  f(x) = 1 + \frac{1}{2!} x + \frac{1}{3!}x^2 + \cdots.
\end{equation}
For $x>0$, every term in the series is positive, leaving no possibility of cancellation error. (The analysis for negative $x$ is more subtle.) If $x\approx 0$, then we should be able to find $n$ such that $x^n/(n+1)!$ is smaller than machine precision, so that adding it to the terms before it will not change the result numerically. Thus a truncated form of the series can serve as a stable method for $x\approx 0$. 

\subsection*{Goals}

You will experiment with the two methods of computing $f$ and observe their relative accuracy. 

\subsection*{Preparation}

Read section 1.3.

\subsection*{Procedure}

Download the template script and complete it to perform the following steps. 

\begin{enumerate}
\item MATLAB has a stable way of computing $y_2$ in~\eqref{direct} without using subtraction. You will use it to get reference ``exact'' values of $f$. Let
\begin{verbatim}
x = logspace(-16,0,500);
\end{verbatim}
  Create a vector \texttt{y} the same size as \texttt{x} such that $y_j$ is the result of \texttt{expm1(x(j))/x(j)}.

\item Create a vector \texttt{z} such that $z_j$ is computed using the three steps in~\eqref{direct} for $x_j$.

\item Compute a vector of relative differences between $z_j$ and $y_j$. Make a log-log plot of the result as a function of $x$. You will see a loss of accuracy as $x\to 0$. 

\item By trial and error, find a value of $n$ such that $0.01^n/(n+1)!$ is less than \texttt{eps} (machine epsilon). This defines the last term to keep in truncating the series~\eqref{series}.

\item Create a vector \texttt{w} such that $w_j$ is computed using the truncated form of~\eqref{series}. Repeat step~3 for \texttt{w} in place of \texttt{z}, using \texttt{semilogx} instead of \texttt{loglog}. This time you should see accuracy maintained as $x\to 0$. (If more terms of the series were kept, the accuracy could be maintained as $x\to 1$ as well, but the direct method is more efficient there.)

\end{enumerate}


\end{document}
