\documentclass[11pt]{article}

\usepackage{amsmath,amsthm}
\usepackage[headings]{fullpage}
\usepackage[utopia]{mathdesign}
\usepackage{color}

\pagestyle{myheadings}
\markboth{Norms}{Norms}

\input{../../fncextra}

\begin{document}

\begin{center}
  \bf Norm-an conquest
\end{center}

The absolute value $|x|$ represents the distance on the number line
between the point $x$ and the origin. We can do the same for a vector
of $n$ numbers, plotting an arrows whose tail is at the origin and
applying the Pythagorean theorem to conclude that $\bfx$ represents a
point that is a distance $(x_1^2 + \cdots + x_n^2)^{1/2}$ from the
origin. This is what we call the 2-norm, $\|\bfx\|_2$. It turns out
that if we abstract the notion of ``distance'' a little to its
essential properties, we can get other vector norms as well. Each norm has its own set of unit vectors (those having norm equal to 1).

For matrices, a definition based on the action of the
matrix as a linear transformation is quite handy for analyzing linear
algebra algorithms.  The definition starts with all the unit vectors
in whatever norm we like. Each of these
vectors $\bfv$ is mapped to the vector $\bfu=\mA\bfv$. The maximum of
$\|\bfu\|$ for all such $\bfv$ is the value of $\|\mA\|$. This
definition can be visualized easily only when both $\bfv$ and
$\bfu$ are low-dimensional.

\subsection*{Goals}

You will use experiments based on the definitions of norms to approximate the calculation of norms of a small matrix.

\subsection*{Preparation}

Read section 2.7. Recall that the equations $x=\cos(\theta)$, $y=\sin(\theta)$, $0\le \theta \le 2\pi$ parameterize the unit circle (set of all unit vectors) defined by the 2-norm in $\mathbf{R}^2$.

\subsection*{Procedure}

Download the script template and complete it to perform the following steps.

\begin{enumerate}
\item Define a vector \texttt{theta} of 200 equally spaced values from 0 to $2\pi$. Use it to define vectors \texttt{x} and \texttt{y} via $x_j = \cos(\theta_j)$, $y_j = \sin(\theta_j)$ (do this without any loops). Plot these points, which lie on the unit circle.

\item Let \texttt{A=magic(2)}. For each $j$, define the vector $\bfv=[x_j\:y_j]^T$ and let $\bfu=\mA\bfv$. On a new graph, plot all the $\bfu$ points, and make a vector storing $\|\bfu\|_2$ for all $j$.

\item Plot $\|\bfu\|_2$ as a function of $\theta$. 

\item Using \texttt{format long} and \texttt{max}, calculate and show the maximum value of $\|\bfu\|_2$ over all $j$. This estimates $\|\mA\|_2$. Also compute the actual value \texttt{norm(A,2)}.
  
\item The script creates new vectors \texttt{x} and \texttt{y} that are unit vectors in the $\infty$-norm. Plot these points, giving a title and axis labels.

\item Repeat step 2, this time collecting $\|\bfu\|_\infty$ as a function of the point index $j$.
  
\item Make a plot as in step 3. This time, the $x$-axis will the index number $j$.

\item Estimate the norm $\|\mA\|_\infty$ and compute its exact value, as in step 4. 
\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
